### 2016-07-01 overview.tex
New pieces of the structure

\label{sec:testbed}

@@

Based on a regular Raspbian Linux image, we explain how
to customize it to include desired packages for performing routinary tasks and set it
in all the \ac{Raspi}s. Also, when working on a testbed with many devices, it may not always be
desired to make persistent changes to the devices. Thus, we later present how to utilize
stacked filesystems to enable both persistent and temporary
storage. This step of the procedure will be optional if the testbed
administrator decides to write the whole image into a memory card. \todot{Check the
veracity and correctness of the previous statement. Question: Shoudln't we use
an overlay FS regardless of how the memory card is configured?}.
Although being simpler than a stacked filesystem, writing the whole Linux
filesystem to several memory cards may complicate future
maintenance of the testbed since updates need to be performed in each \ac{Raspi}.
Even when tools such as Python Fabric may ease this task, the maintenance might still
be cumbersome in some cases. Therefore, we propose and encourage the alternative centralized
approach where the initial \ac{RAM} filesystem is customized in a way that
enables \ac{Raspi}s to fetch the its root filesystem from a \ac{HTTP} server as
mentioned previously. The following procedure explains how the testbed may be setup and configured.

@@

\subsubsection{Installing Raspbian}

To get started, we first need to install an \ac{OS} on the
\ac{Raspi}s. We will use the popular Debian-based Raspbian
Linux~\cite{raspbian}

We will be using a \ac{PC} running a Debian-based Linux distribution
to setup a customized Raspbian Linux image that can be written to
a memory card and work in any \ac{Raspi} version 1 or 2.

Raspbian is made available in two bundles: Raspbian and Raspbian lite.
The difference between the two is that Raspbian contains a pre-installed desktop
environment, and Raspbian lite is by default only accessible through a 
command shell. Both Raspbian and Raspbian lite will both by default be remotely
accessible using \ac{SSH}. \ac{SSH} provides the means to remotely
connect to another machine through a command shell.
%In fact it is the successor
%of \ac{Telnet} that has been designed with network security in mind.

The \ac{Raspi}s in our testbed are not connected to monitors, so we
have decided to go with Raspbian lite.
%Our testbed does not have monitors so we will install Raspbian lite.
A desktop environment can be installed using the package manager
(apt-get) in case a desktop environment will be desired at a later
time.

Below is the procedure we followed to setup the \ac{Raspi}s in our testbed. We
have performed a simple setup where each \ac{Raspi} is equipped with a
memory card that contains a slightly customized Raspbian lite image. The image
of the \ac{OS} in all \ac{Raspi}s will be identical.

The overall procedure to customize the official Raspbian lite image
are:
\begin{enumerate}
    \item Download Raspbian lite
    \item Alter Raspbian lite. e.g. browse, modify, add, and delete files
        in the official Raspbian lite image
    \item Change root, i.e. change root filesystem into the Raspbian lite
        image to update and install software packages
    \item Write image to memory cards
\end{enumerate}

Hashmark (\#) and dollar (\$) will be used throughout the code blocks in
the paper to indicate if an operation needs to be run with root permissions or
user permissions.

\subsection{Download Raspbian lite}

The latest Raspbian lite bundle can be downloaded from Raspbian's official
webpage~\cite{raspbian}.
%To download the latest Raspbian lite, we go to the url
%\url{http://downloads.raspberrypi.org/raspbian_lite/images/}.
At the time of this writing, the latest available bundle was
\texttt{2016-05-27-raspbian-jessie-lite.zip}.

To ensure that the content of the bundle does not change, this procedure
will be based on that particular version of Raspbian lite which we have
made available at~\cite{tunescode_webpage}. All other files used in this paper
will also be available there unless specified otherwise.

%We want to proceed downloading the bundle in a shell instead of the browser.
%To do this, first open a command shell on a linux machine. This guide is
%assuming a Debian-based distribution, but there should only be minor differences
%if you have another package manager. i.e. the package managers and name of software packages.

To get started, open a Linux shell (terminal) and declare the variables below.
The \texttt{URL} and \texttt{IMAGE} specifies where the Linux bundle will be
located and \texttt{WORKDIR} specifies a working directory where we will 
store Raspbian and customize it.
Do not type in \$ and \# in the shell. Those indicates if a command can be
executed as a regular user or if root permissions are required.
Running a command as root can on most systems be done by putting
\texttt{sudo} in front of the command. This is illustrated in the
following code block 
with the command \texttt{whoami} that prints the username. Lines within a code
block without leading \$ or \# is terminal output or content of a file.

%After the terminal has been opened, we start by declaring a few variables to reduce
%repeated typing. The first variables we declare is the Raspbian image name.
%Notice that the extension has been omitted. This is because the image has been
%packed into a zip file. The other variable we declare is a working directory.
%This is where we will download the image to and work on it. In other words, it
%will be stored in \texttt{/home/<username>/Raspbian}

%To download Raspbian lite, we go to the \url{http://downloads.raspberrypi.org}.
%There should be a folder called raspbian\lite and
%to find the download page at which Raspbian lite should be located. Instead of
%downloading with the browser, we just extract the download url to the latest
%image. At the time of this writing, that was

% Define variable]
\begin{lstlisting}[]
$ export URL="http://kom.aau.dk/project/TuneSCode/raspi/"
$ export IMAGE="2016-05-27-raspbian-jessie-lite"
$ export WORKDIR="${HOME}/Raspbian"
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

%A final thing to notice is the way \$ and \# are used in the code blocks. \$
%means that the command can be run with normal user permissions where \# indicates that the
%command needs root permissions.
%Running a command with root permissions can on most systems be done by putting
%the command "sudo" in front of each line. This is illustrated in the code below
%with the command "whoami" that prints the username of the caller. Lines without
%leading \$ or \# is the terminal output.

% Root example
\begin{lstlisting}[]
$ whoami
<USERNAME>
$ sudo whoami
root
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Next, make the working directory and used \texttt{cd} to change into the working
directory. Then \texttt{wget} can be used to download Raspbian lite before
unpacking the \texttt{zip} file:

% Download and unpack image
\begin{lstlisting}[]
$ mkdir -p ${WORKDIR}
$ cd ${WORKDIR}
$ wget ${URL%/*}/${IMAGE}.zip
$ unzip ${IMAGE}.zip
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

%A more recent version of Raspbian lite may be available at~\cite{raspbian}.
%\url{http://downloads.raspberrypi.org/raspbian_lite/images}.

\subsection{The image}

After Raspbian lite has been unpacked, there should be an \texttt{img} file in the directory.
\texttt{fdisk} can be used to display the content of the image. Parse the arguments
"-u sectors" to display the sizes in sectors and "-l" to display the partitions within the
image and and automatically exit \texttt{fdisk} again:

% Check out the image   % The dollar hack was to fix vim syntax
\begin{lstlisting}[literate={DOLLAR}{\$}1]
DOLLAR fdisk -u sectors -l ${IMAGE}.img
Disk 2016-05-27-raspbian-jessie-lite.img: 1.3 GiB, 1387266048 bytes, 2709504 sectors
Units: sectors of 1 * 512 = 512 bytes
Sector size (logical/physical): 512 bytes / 512 bytes
I/O size (minimum/optimal): 512 bytes / 512 bytes
Disklabel type: dos
Disk identifier: 0x6fcf21f3

Device                               Boot  Start     End Sectors  Size Id Type
2016-05-27-raspbian-jessie-lite.img1        8192  137215  129024   63M  c W95 FAT32 (LBA)
2016-05-27-raspbian-jessie-lite.img2      137216 2709503 2572288  1.2G 83 Linux
\end{lstlisting}
\FloatBarrier

The output provides knowledge about the image. The image is in total
2709504 sectors (1.3GiB) in size and contains two
partitions. The first partition starts at sector 8192 and
the other partition starts at sector 137216. The partitions is of type FAT32
and Linux and are 63M and 1.2G big. This indicates that the first partition
is most likely a boot partition and the second partition is a traditional Linux file
system. Actually, it is the Linux root filesystem (i.e. /).

\subsection{Resize image}
1.2G is not very much to store the root filesystem given that it is desired to
alter the files and install additional packages. It may therefore be
necessary to increase the partition size.
The following illustrates how the image and its root filesystem
can be expanded by one \ac{GB}.

First, expand the image one \ac{GB} and wait for the data to be completely written
to the disk:
\begin{lstlisting}[]
$ dd if=/dev/zero bs=1M count=1024 >> ${IMAGE}.img && sync
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Use \texttt{fdisk} to see that the image is now one \ac{GB} larger:
\begin{lstlisting}[literate={DOLLAR}{\$}1]
DOLLAR fdisk -u sectors -l ${IMAGE}.img
Disk 2016-05-27-raspbian-jessie-lite.img: 2.3 GiB, 2461007872 bytes, 4806656 sectors
Units: sectors of 1 * 512 = 512 bytes
Sector size (logical/physical): 512 bytes / 512 bytes
I/O size (minimum/optimal): 512 bytes / 512 bytes
Disklabel type: dos
Disk identifier: 0x6fcf21f3

Device                               Boot  Start     End Sectors  Size Id Type
2016-05-27-raspbian-jessie-lite.img1        8192  137215  129024   63M  c W95 FAT32 (LBA)
2016-05-27-raspbian-jessie-lite.img2      137216 2709503 2572288  1.2G 83 Linux
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Now, to expand the root filesystem it is required to first remove the Linux partition
and then add it again one \ac{GB} bigger.
Use \texttt{fdisk} to alter the partition table. The \texttt{echo}'s within the parenthesis
will be interpreted as
key-presses in \texttt{fdisk} and hence delete the last partition, create a new
primary partition, print new partition table, and write new partition table
to the image:

\begin{lstlisting}[]
$ (echo d; echo ; echo n; echo p; echo ; echo 137216; echo ; echo p; echo w) | fdisk ${IMAGE}.img
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Use a loopback device to make the Raspbian image available as a block device: 
\begin{lstlisting}[]
# export DEV=$(sudo losetup --show -f -P ${IMAGE}.img); echo $DEV
/dev/loop0
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

\texttt{lsblk} may be used to list block devices. It appears that the image is
available as \texttt{/dev/loop0} which has two partitions \texttt{loop0p1} and
\texttt{loop0p2}:
%Use \texttt{lsblk} to view the partitions:

\begin{lstlisting}[]
# lsblk
NAME      MAJ:MIN RM  SIZE RO TYPE MOUNTPOINT
...
loop0       7:0    0  2.3G  0 loop 
|-loop0p1 259:2    0   63M  0 loop 
|-loop0p2 259:3    0  2.2G  0 loop
...
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Check and resize the root partition:
\begin{lstlisting}[]
# e2fsck -f ${DEV}p2
# resize2fs ${DEV}p2
\end{lstlisting}
\FloatBarrier

%We see that the boot and root partitions starts at sector 8192 and 137216 respectivly.
%We also notice that each sector is 512 bytes.
%On newer systems, we can mount the device easier using "losetup --show -f -P IMAGE"
%Mount root partition:

\subsection{Mount image}

Browsing and altering the files in the image is easy. Simply mount the partitions.
%To browser and alter the files in the image, it is possible to mount the partitions.

Lets first mount the root partition. This is done by creating an empty directory
that we will be used as a mountpoint. We name it \texttt{root} and create it in the
working directory before mounting the root filesystem onto the mountpoint.
%After this, mount the partitions:
%After this,
%we mount the partition starting at sector 137216. To provide this offset information
%to mount, we need to find the offset in bytes. From the fdisk command we could also
%see that each sector is 512 bytes. Thus, we multiply the sector size and the sector
%start to get the offset in bytes.

\begin{lstlisting}[]
$ export ROOTDIR="${WORKDIR}/root"
$ mkdir -p ${ROOTDIR}
# mount ${DEV}p2 ${ROOTDIR}
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}
%# mount -o loop,offset=$((137216*512)) ${IMAGE}.img ${ROOTDIR}

The root filesystem already has a boot directory that can be used as
the mountpoint for the boot partition.
%After this, we can also mount the boot partition inside the newly mounted
%root filesystem.
This is convenient because that partition will be mounted
on this directory when a \ac{Raspi} starts up with a memory card containing
the image. %It is therefore the natural place to mount it for later purposes.
Now, mount boot partition:
\begin{lstlisting}[]
# mount ${DEV}p1 ${ROOTDIR}/boot
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}
%# mount -o loop,offset=$((8192*512)) ${IMAGE}.img ${ROOTDIR}/boot

%We can now change all the files in the disk image as desired.
It is now possible to change all files within the Raspbian image as desired.
We will take advantage of this to edit configuration files, append new files,
and even update and install packages.

\subsection{Configuring OS files and scripts}

The \ac{Raspi}s should be setup as similar as possible. One of those
few settings that may be different in the devices is their hostname.
%The hostname helps the user to distinguish the devices from each other.
%The hostname helps distinguish them from each other.
One approach to create unique hostnames would be to write the hostname
in each memory card. Another way is to create a start up script to automatically
configure the hostname. We decided to define the hostnames based on the \ac{MAC}
addresses of the \ac{Raspi}s wired ethernet interface. We store the \ac{MAC}
addresses and hostnames of the \ac{Raspi}s in a file as below. The testbed
administrator will need to insert the \ac{MAC} addresses and hostnames of
his/her \ac{Raspi}s.

%The \ac{Raspi}s should be setup to be as similar as possible. One of those
%things that is preferred to be different on the devices is their hostname.
%This helps distinguish them from each other. One solution to this in our
%approach would be to uniquely change the hostname in each memory card.
%This is however not desired because that would either require modifying
%the customized Raspbian lite image before writing the content to the
%memory card or to manually change it in the \ac{Raspi}s after the image has
%been written to a memory card.
%Instead, we decided to register the ethernet MAC addresses of all the \ac{Raspi}s
%in our testbed. Those had to be used to put them in a domain name anyway.
%
%Below is a part of the file where we keep track of the MAC addresses and assign
%hostnames. This file will be available online and copied to the Raspbian lite
%image in a later step under the listed filename. For different \ac{Raspi}s
%the MAC addresses has to be changed. The MAC address of a network card can be found
%using the command \texttt{"ifconfig"} or \texttt{"ip addr"}.

%Instead, we decided to register the ethernet MAC addresses of all the \ac{Raspi}s
%in our testbed. Those had to be used to put them in a domain name anyway.

%Below is a snippet of the file of all the hostnames and \ac{MAC} addresses:

%Below is a part of the file where we keep track of the MAC addresses and assign
%hostnames.
%This file will be available online and copied to the Raspbian lite
%image in a later step under the listed filename. For different \ac{Raspi}s
%the MAC addresses has to be changed. The MAC address of a network card can be found
%using the command \texttt{"ifconfig"} or \texttt{"ip addr"}.

% MAC and Hostname file
\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/home/pi/rasp\_config/nodes.csv}@>
<@\textcolor{gray}{
---------------------------------------------------------------}
\Reactivatenumber @>
# Ethernet MAC    Hostname
b8:27:eb:5b:da:20 rasp00
b8:27:eb:7b:c3:91 rasp01
b8:27:eb:54:9c:64 rasp02
b8:27:eb:95:bd:11 rasp03
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

%Provided that each \ac{Raspi} will have knowledge of all other \ac{Raspi}s in
%the testbed, it still needs to know which name to give itself.
%We do this using a small Bash script:
%We use the \ac{Bash} script below to enable each \ac{Raspi} to find its
%hostname in \texttt{nodes.csv}.
A \ac{Bash} script can then be used to have each \ac{Raspi} assign its hostname:

% Set hostname
\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/home/pi/rasp\_config/set\_hostname}@>
<@\textcolor{gray}{
---------------------------------------------------------------}
\Reactivatenumber @>
#!/usr/bin/env bash

script_path="$(dirname $(realpath $0))"
config_file=${script_path}/nodes.csv 
mac=$(cat /sys/class/net/eth0/address)
old_hostname=$(hostname)
new_hostname=$(grep $mac $config_file | cut -f2 -d' ')

# Assign hostname found in nodes.csv
if [ ! -z ${new_hostname} ]; then
    echo ${new_hostname} > /etc/hostname
    hostname ${new_hostname}
    sed -i.old -e "s:${old_hostname}:${new_hostname}:g" /etc/hosts
fi
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

The code lines above do the following:
\begin{itemize}
    \item Line 1: Tells system to interpret the script using \ac{Bash}
    \item Line 3-4: Gets the path to the script itself and the list of hostnames
    \item Line 5: Gets the \ac{MAC} address of the node itself
    \item Line 6: Gets the current hostname
    \item Line 7: Gets the new hostname from the hostname list
    \item Line 10-14: Assigns the new hostname
\end{itemize}

%Line 1 tells the system to interpret the script using Bash. Line 3 gets the
%path of the script itself. Line 4 gets the MAC address. This script will be
%located in all \ac{Raspi}s and thus return the MAC address of the node itself
%when executed.
%Line 4 searches line by line for this mac address.
%Line 8-11 assigns the hostname found in \texttt{"nodes.csv"}. If the MAC did not exists,
%then it will keep the default hostname of the system.

It is likely that the testbed administrator needs to either change or add
scripts to configure the \ac{Raspi}s. This would be tedious to distribute to
all \ac{Raspi}s at a later stage. We ease this task by making the \ac{Raspi}s fetch the
configuration scripts during start up before running the configuration scripts.

The below script attempts to download/update the scripts:

\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/home/pi/rasp\_config/update\_rasp\_config}@>
<@\textcolor{gray}{
---------------------------------------------------------------}
\Reactivatenumber @>
#!/usr/bin/env bash

url="http://kom.aau.dk/project/TuneSCode/raspi/"
config_file="rasp_config.zip"

# Attempt to fetch new configuration files
if ! wget -q --show-progress -O /tmp/${config_file} ${url%/}/${config_file}; then
    echo "Warning: Unable to update rasp_config files"
    exit 1
fi

# Unzip and overwrite configurationn files to root's home directory
unzip -q -o /tmp/${config_file} -d /home/pi/
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

\begin{itemize}
    \item Line 3-4: Is the url and zip file that should be downloaded
    \item Line 7-10: Downloads the new configuration files to \texttt{/tmp}
        and prints a warning in case of errors
    \item Line 13: Unzips the files to \texttt{/home/pi/} while ignoring existing files/directories
\end{itemize}

We have packed the files and uploaded them to a webserver. These can be
retrieved with \texttt{wget} and unzipped into the Raspbian lite image:

% Get the files
\begin{lstlisting}[]
$ wget ${URL%/*}/rasp_config.zip
$ unzip rasp_config.zip -d ${ROOTDIR}/home/pi/
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

For the above scripts to work in the \ac{Raspi}s, it is required that the \ac{Raspi}
\ac{MAC} address is found in \texttt{nodes.csv}. Thus,
\texttt{rasp\_config.csv} needs to be unzipped, altered, and zipped again in
the \ac{HTTP} server from which the \ac{Raspi} retrieves the files from.
The testbed administrator will therefore need to also edit the \ac{URL} in the
script \texttt{update\_rasp\_config}.
%"nodes.csv" can be changed using a text editor. For vim, this could be done
%the following way:
%\begin{lstlisting}[]
%$ vim ${ROOTDIR}/home/pi/rasp_config/nodes.csv
%\end{lstlisting}
%\FloatBarrier
%\vspace{-5mm}

Finally, to actually make the \ac{Raspi}s change hostname, we have to make
each \ac{Raspi} call the above script when it starts up. Optionally,
also run the update script before running other scripts.
Do this with root permissions by inserting the line(s) below before
\texttt{"exit 0"} using your favorite editor:
%"\$\{ROOTDIR\}/etc/rc.local" using
%your favorite text editor..
%The file
%can be opened with your favorite text editor.

\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/etc/rc.local}@>
<@\textcolor{gray}{
-----------------------------------------------------}
\Reactivatenumber @>
...
bash /home/pi/rasp_config/update_rasp_config
bash /home/pi/rasp_config/set_hostname
...
exit 0
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Configuring \texttt{rc.local} makes it really cumbersome for the testbed
administrator to add new scripts since it first needs to be included
in \texttt{rasp\_config.zip} and then each \ac{Raspi} needs to have an additional 
line added to \texttt{rc.local} in the root image or in all memory cards depending on
the final testbed setup.

A solution to this is to make yet another script in \texttt{rasp\_config.zip}
that calls all the other scripts in the desired order. This would look like
below:
\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/home/pi/rasp\_config/main}@>
<@\textcolor{gray}{
-----------------------------------------------------}
\Reactivatenumber @>
#!/usr/bin/env bash

bash /home/pi/rasp_config/set_hostname
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Now, to use the above script, \texttt{rc.local} needs to look like this:
\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/etc/rc.local}@>
<@\textcolor{gray}{
-----------------------------------------------------}
\Reactivatenumber @>
...
bash /home/pi/rasp_config/update_rasp_config
bash /home/pi/rasp_config/main
...
exit 0
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}
Notice that \texttt{set\_hostname} is now called by the \texttt{main}
script instead. The update script is still called directly. This ensures that
all configuration scripts are updated before executed. Changes to the update
script itself will first take effect at next system startup.
%Had the update script
%been called in \texttt{main} instead would mean that changes to the 
%\texttt{main} script would also first have been available at next startup.

%This can be done by inserting a call to the script
%set\_hostname in rc.local before it exits.
%
%Call set\_hostname script at startup. We insert a line of code to call script in rc.local just after the initial comments (i.e. lines starting with \#).
%\begin{lstlisting}[]
%# line_number=$(egrep -n -m1 "(^[^#])|(^$)" ${ROOTDIR}/etc/rc.local | cut -d: -f1)
%# sed -i "${line_number}c bash /home/pi/rasp_config/set_hostname" ${ROOTDIR}/etc/rc.local
%\end{lstlisting}
%\FloatBarrier
%$ line="bash /home/pi/.rasp_config/set_hostname"
%# awk -v text="$line" '!/^#/ && !p {print text; p=1} 1' ${ROOTDIR}/etc/rc.local > <@ \Suppressnumber @>
%    ${ROOTDIR}/etc/rc.local <@ \Reactivatenumber @>

%WE NEED TO ADD A FEW COMMANDS TELLING HOW TO ZIP, MODIFY, AND UNZIP
%FOR THE READER TO BE ABLE TO UPLOAD A CONSTRUCT HIS/HER OWN BUNDLE.

%All other files could be updated as well.
It is not only possible to add and configure files to/within the image. One can
also install and update the packages within the image which will be illustrated
in the next section.

\subsection{Installing and updating the image}

It may be desired to pre-install some programs in the image before it is
written to all the memory cards that goes into the \ac{Raspi}s.
This can be done from any linux x86 machine using QEMU Chroot~\cite{QemuUserEmulation}
(change root).
%\url{https://wiki.debian.org/QemuUserEmulation}
%\url{https://wiki.archlinux.org/index.php/Raspberry_Pi}

Due to the \ac{ARM} processor that \ac{Raspi}s are using, it is required to
install some additional software:
%Because \ac{Raspi}s are equipped with an \ac{ARM} processor 
%%it is not as
%straightforward as performing the task for two Linux systems of the same
%architecture.

% CHROOT to OS image
\begin{lstlisting}[]
# apt-get install binfmt-support qemu qemu-user-static
# update-binfmts --display qemu-arm
qemu-arm (enabled):
     package = qemu-user-static
        type = magic
      offset = 0
       magic = \x7fELF\x01\x01\x01\x00\x00\x00\x00\x00\x00\x00\x00\x00\x02\x00\x28\x00
        mask = \xff\xff\xff\xff\xff\xff\xff\x00\xff\xff\xff\xff\xff\xff\xff\xff\xfe\xff\xff\xff
 interpreter = /usr/bin/qemu-arm-static
    detector = 
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}
% This was neccessary on arch
%# update-binfmts --importdir /var/lib/binfmts/ --import

Make sure the second command writes "enabled" as the output above. If that 
is not the case, then try enabling it:

\begin{lstlisting}[]
# update-binfmts --enable qemu-arm
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Provided that qemu-arm is enabled, we should now be able to change root (chroot)
into our Raspbian lite image. Change root is a method in Linux that enables
the root (/) to be changed. Thus, enabling a Linux installation within a
Linux installation.

There are a few commands to be performed before actually changing root into the
root partion of the image.
First, to get internet access from within the Raspbian lite image it is needed
to copy our \texttt{resolv.conf} into the filesystem. Line 1 change directory
into root filesystem of the image. The second line copies \texttt{resolv.conf}
from the testbed administrators \ac{PC} to the image.

% CHROOT to OS image
\begin{lstlisting}[]
$ cd $ROOTDIR
# cp /etc/resolv.conf ${ROOTDIR}/etc/resolv.conf
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Now, because of the \ac{ARM} architecture,
\texttt{/usr/bin/qemu-arm-static} needs to be copied into the image before continuing:

% CHROOT to OS image
\begin{lstlisting}[]
# cp /usr/bin/qemu-arm-static ${ROOTDIR}/usr/bin
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

The final preparation before changing root is to populate the directories \texttt{proc},
\texttt{sys}, and \texttt{dev}:

% CHROOT to OS image
\begin{lstlisting}[]
# mount  -t proc proc proc/
# mount --bind /sys sys/
# mount --bind /dev dev/
# mount --bind /dev/pts dev/pts
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Finally, it is time to change root.
%Finally, the filesystem is ready for us to change root.
This can be done with \texttt{proot}. It may be required to install \texttt{proot}
using "\texttt{apt-get install proot}":

% CHROOT to OS image
\begin{lstlisting}[]
# proot -q qemu-arm-static -S ${ROOTDIR}
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Most available material online uses the more known \texttt{chroot} command as
written in the code block below instead of \texttt{proot}. This did not work
correctly in our machines, but we present the command as an alternative to
\texttt{proot} in case the testbed administrator should be in the opposite
situration where \texttt{proot} is not working.

% CHROOT to OS image
\begin{lstlisting}[]
# chroot ${ROOTDIR} /usr/bin/qemu-arm-static /bin/bash    (ALTERNATIVE TO proot)
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

If everything went well, we should now be \texttt{chrooted} into the Raspbian lite
filesystem under its root. Optionally, change the prompt title to indicate that it
is a \texttt{chrooted} environment:

% Optionally, we may create a unique prompt to indicate we have changed root
\begin{lstlisting}[]
# export PS1="(chroot) $PS1"
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

The Raspbian lite system image should now be possible to use almost as if it
had been booted in a \ac{Raspi}.
A major difference is that the testbed administrator's \ac{PC} is likely
significantly faster than a \ac{Raspi}.
%The only difference is that the laptop that we are using is
%significantly faster than a \ac{Raspi}.
Thus enabling updates, upgrades, and installing new software packages much
faster than in the \ac{Raspi}s.
%Thus enabling us to for example update,
%upgrade, and install software packages:

% Update system
\begin{lstlisting}[]
(chroot) # apt-get update
(chroot) # apt-get upgrade
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

We also install a few packages that we find useful:
%Lets install some useful applications:
% Install packages
\begin{lstlisting}[]
(chroot) # apt-get install vim git screen
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

When writing the image to a memory card, all these change that has been made
to the image so far will exists in all \ac{Raspi}s.
%All the changes that are made here will exists in all \ac{Raspi}s when the
%image is written to a memory card.

\subsection{Installation an overlaying filesystem}

Our goal was to enable the users of the testbed to change however the
\ac{Raspi} are configured while still being able to revert those changes easily.

We do this by storing a users work in the \ac{Raspi}s in \ac{RAM}. Thus a
\ac{Raspi}, when a \ac{Raspi} is rebooted, it will forget all changes.

A way to do this is using stacked filesystem. Stacking filesystem enables
a filesystem to overlay another filesystem. Whenever a file is accessed,
the upper filesystem will forward the request to the lower filesystem in
case it does not have it itself. If it has it, it will simpy return the file.

This idea can be used in our setup to mount the root filesystem (i.e. Raspbian
lite) in the \ac{Raspi}s duing startup as read-only. Thus, the files will not
be changable. Some programs needs to change files and the testbed admin might
want to change configurations. To enable this, it is possible to
overlay a filesystem that is mounted in \ac{RAM} as read/writable on top of
the regular root. Thus, reading a file may return a file from the lower filesytem,
but if it is stored, it will be saved in the upper filesystem. Accessing this file
again will return the stored file from the upper layer.

%We do this using an overlay filesystem. The idea is to let the \ac{Raspi}s
%mount their root filesystem as 
%
%We do this using an overlay filesystem (overlayfs).
%Overlayfs enables a filesystem to be stacked on top of another filesystem. Thus,
%if the \ac{Raspi}s mount their root filesystem (/) as read-onlyif we mount our root (/) filesystem as 
%
%
%Overlayfs enables filesystems
%to be stacked on top of each other. We will be using this to make a lower
%filesystem and an upper filesystem. The lower filesystem will be our
%traditional filesystem and the upper filesystem will be mounted in tmpfs
%(i.e. mounted in \ac{RAM}).
%
%Thus, when a file is requsted from stacked filesystem, it will first attempt
%to fetch the file in the upper filesystem if it does not exists there, it will
%request it from the lower filessytem.

Assuming that you are still \texttt{chrooted} into the Raspbian lite filesystem, we
can setup the overlay filesystem (overlayfs).

There already exists implementations overlaying the root filesystem. We will
use an existing implementation
available at github: \cite{overlayroot}.

We installed git in a previous step, so lets clone the repository. The below
code stores it in /tmp which is really mounted in \ac{RAM}. Thus, the files
will disappear when the machine is rebooted.

% Get files
\begin{lstlisting}[]
(chroot) # OVERLAYROOTDIR="/tmp/overlayroot"
(chroot) # git clone https://github.com/chesty/overlayroot.git $OVERLAYROOTDIR
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

To enable the overlaying filesystem it is required to generate an initial \ac{RAM}
file system (initramfs). Initramfs is an initial filesystem that gets loaded into
\ac{RAM} during the startup process of a linux machine to prepare the real
filesystem.

For this purpose, we need BuxyBox:

% Install required packages
\begin{lstlisting}[]
(chroot) # apt-get install busybox
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Add and activate the overlaying filesystem, we need to add the scripts and tell initramfs
to load the overlaying filesytem:

% Setup initramfs
\begin{lstlisting}[]
(chroot) # cp ${OVERLAYROOTDIR}/hooks-overlay /etc/initramfs-tools/hooks/
(chroot) # cp ${OVERLAYROOTDIR}/init-bottom-overlay /etc/initramfs-tools/scripts/init-bottom/
(chroot) # echo "overlay" > /etc/initramfs-tools/modules
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

%Finally, lets generate the initial \ac{RAM} filesystem. This is done using the command
%\texttt{mkinitramfs}, that will by default look for the kernel modules of the current system. Remember we are
%chrooted, so we need to specify the correct kernel modules to look for. These
%can be found in "/lib/modules".
Now, the initial \ac{RAM} filesystem can be generated. \ac{Raspi} version 1
needs a different kernel than \ac{Raspi} version 2 and version 3. Thus, start
by finding the available kernel modules:
%Because \texttt{initramfs} is executed within a
%\texttt{chrooted} environment, it will not be able to detect the
%kernel modules in the \texttt{chrooted} environment. The kernel modules will

% Create an initramfs
\begin{lstlisting}[]
(chroot) # ls /lib/modules/
4.4.13+  4.4.13-v7+
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Kernel "4.4.13+" is for \ac{Raspi} version 1 and kernel "4.4.13-v7+" works
for \ac{Raspi} version 2 and version 3.

Lets proceed by generating an initial \ac{RAM} filesystem for both kernels:

%Lets proceed with kernel modules "4.4.13+".
%Lets create an initial ram filesystem for both kernels.

%Thus, to run mkinitramfs, type in the correct kernel version:

% Create an initramfs
\begin{lstlisting}[]
(chroot) # mkinitramfs -o /boot/init.gz -k 4.4.13+
(chroot) # mkinitramfs -o /boot/init-v7.gz -k 4.4.13-v7+
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

%IT SEEM THAT THE RASPIS DEMANDS DIFFERENT KERNEL MODULES AFTER ALL.
%RASP-VERSION1=4.4.13+, RASP-VERSION2=4.4.13-V7+, RASP-VERSION3=4.4.13-V7+

Although there might be printet a few warnings, the above commands should successfully
generate working initial \ac{RAM} filesystems.

An initial \ac{RAM} filesystem needs to be called by the bootloader. In Raspbian
this is done by adding a command to \texttt{config.txt} in the boot partition.

>>>>

\label{sec:overview}

@@

Based on a regular Raspbian Linux image, we explain how
to customize it to include desired packages for performing routinary tasks and set it
in all the \ac{Raspi}s. Also, when working on a testbed with many devices, it may not always be
desired to make persistent changes to the devices. Thus, we later present how to utilize
stacked filesystems to enable both persistent and temporary
storage. This step of the procedure will be optional if the testbed
administrator decides to write the whole image into a memory card. \todot{Check the
veracity and correctness of the previous statement. Question: Shoudln't we use
an overlay FS regardless of how the memory card is configured?}.
Although being simpler than a stacked filesystem, writing the whole Linux
filesystem to several memory cards may complicate future
maintenance of the testbed since updates need to be performed in each \ac{Raspi}.
Even when tools such as Python Fabric may ease this task, the maintenance might still
be cumbersome in some cases. Therefore, we propose and encourage the alternative centralized
approach where the initial \ac{RAM} filesystem is customized in a way that
enables \ac{Raspi}s to fetch the its root filesystem from a \ac{HTTP} server as
mentioned previously.

%The following procedure explains how the testbed may be setup and configured.

@@
